---
title: "Phylogenetic Diversity - Traits"
author: "Student Name; Z620: Quantitative Biodiversity, Indiana University"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW
Up to this point, we have been focusing on patterns taxonomic diversity in Quantitative Biodiversity. 
Although taxonomic diversity is an important dimension of biodiversity, it is often necessary to consider the evolutionary history or relatedness of species. 
The goal of this exercise is to introduce basic concepts of phylogenetic diversity. 

After completing this exercise you will be able to:

1.  create phylogenetic trees to view evolutionary relationships from sequence data
2.  map functional traits onto phylogenetic trees to visualize the distribution of traits with respect to evolutionary history
3.  test for phylogenetic signal within trait distributions and trait-based patterns of biodiversity

## Directions:
1. Change "Student Name" on line 3 (above) with your name.
2. Complete as much of the exercise as possible during class; what you do not complete in class will need to be done on your own outside of class.
3. Use the handout as a guide; it contains a more complete description of data sets along with the proper scripting needed to carry out the exercise.
4. Be sure to **answer the questions** in this exercise document; they also correspond to the handout.
Space for your answer is provided in this document and indicated by the ">" character.
If you need a second paragraph be sure to start the first line with ">".
5. Before you leave the classroom, **push** this file to your GitHub repo.
6. For homework, follow the directions at the bottom of this file. 
7. When you are done, **Knit** the text and code into a PDF file.
8. After Knitting, please submit the completed exercise by creating a **pull request** via GitHub.
Your pull request should include this file *PhyloTraits_exercise.Rmd* and the PDF output of `Knitr` (*PhyloTraits_exercise.pdf*).


## 1) SETUP

Typically, the first thing you will do in either an R script or an RMarkdown file is setup your environment. 
This includes things such as setting the working directory and loading any packages that you will need.

In the R code chunk below, provide the code to:  
1. clear your R environment,  
2. print your current working directory,  
3. set your working directory to your "*/Week6-PhyloTraits*" folder, and  
4. load all of the required R packages (be sure to install if needed).  

```{r}


```

## 2) DESCRIPTION OF DATA

The maintenance of biodiversity is thought to be influenced by **trade-offs** among species in certain functional traits. 
One such trade-off involves the ability of a highly specialized species to perform exceptionally well on a particular resource compared to the performance of a generalist. 
In this exercise, we will take a phylogenetic approach to mapping phosphorus resource use onto a phylogenetic tree while testing for specialist-generalist trade-offs. 


## 3) SEQUENCE ALIGNMENT


***Question 1***: Using nano or your favorite text editor, compare the `p.isolates.fasta` file and the `p.isolates.afa` file. 
Describe the differences that you observe between the files. 

> ***Answer 1***: 

In the R code chunk below, do the following:
1. read your alignment file,
2. convert the alignment to a DNAbin object,
3. select a region of the gene to visualize (try various regions), and
4. plot the alignment using a grid to visualize rows of sequences.


```{r}


```


***Question 2***:  Make some observations about the `muscle` alignment of the 16S rRNA gene sequences for our bacterial isolates and the outgroup, *Methanosarcina*, a member of the domain archaea. 
Move along the alignment by changing the values in the `window` object. 

a. Approximatley how long are our reads?  
b. What regions do you think would are appropriate for phylogenetic inference and why?  

> ***Answer 2a***:   
> ***Answer 2b***:  


## 4) MAKING A PHYLOGENETIC TREE

Once you have aligned your sequences, the next step is to construct a phylogenetic tree.
Not only is a phylogenetic tree effective for visualizing the evolutionary relationship among taxa, but as you will see later, the information that goes into a phylogenetic tree is needed for downstream analysis. 


### A. Neighbor Joining Trees

In the R code chunk below, do the following:  
1. calcualte the distance matrix using `model = "raw"`,  
2. create a Neighbor Joining tree based on these distances,  
3. define "Methanosarcina" as the outgroup and root the tree, and  
4. plot the rooted tree. 

```{r}


```

***Question 3***: What are the advantages and disadvantages of making a neighbor joining tree?   
**<-- This seems like an important question, but have we provided them with enough resources or direction to be able to answer this question?**


> ***Answer 3***:  
 


### B) SUBSTITUTION MODELS OF DNA EVOLUTION 

In the R code chunk below, do the following:  
1. make a second distance matrix based on the Felsenstein 84 substitution model,  
2. create a saturation plot to compare the *raw* and *Felsenstein (F84)* substitution models,  
3. make Neighbor Joining trees for both, and  
4. create a cophylogenetic plot to compare the topologies of the trees.

```{r}


```


In the R code chunk below, do the following:  
1. pick another substitution model,  
2. create and distance matrix and tree for this model,  
3. make a saturation plot that compares that model to the *Felsenstein (F84)* model,  
4. make a cophylogenetic plot that compares the topologies of both models, and  
5. be sure to format, add appropriate labels, and customize each plot.

```{r}


```


***Question 4***:  

a. Describe the substitution model that you chose. What assumptions does it make and how does it compare to the F84 model?
b. Using the saturation plot and cophylogenetic plots from above, describe how your choice of substitution model affects your phylogenetic reconstruction. 
If the plots are inconsistent with one another, explain why.
c. How does your model compare to the *F84* model and what does this tell you about the substitution rates of nucleotide transitions?


> ***Answer 4a***:   
> ***Answer 4b***:   
> ***Answer 4c***:   

### C) ANALYZING A MAXIMUM LIKELIHOOD TREE

In the R code chunk below, do the following:  
1. Read in the maximum likelihood phylogenetic tree used in the handout.
2. Plot bootstrap support values onto the tree


```{r}


```

***Question 5***:  

a) How does the maximum likelihood tree compare the to the neighbor-joining tree in the handout? 
If the plots seem to be inconsistent with one another, explain what it giving rise to the differences.

b) Why do we bootstrap our tree?

c) What do the bootstrap values tell you? Which branches have very low support? Should we trust these branches? **<-- This is three questions**

> ***Answer 5a***:   
> ***Answer 5b***:   
> ***Answer 5c***:   


## 5) INTEGRATING TRAITS AND PHYLOGENY
### A. Loading Trait Database

In the R code chunk below, do the following:  
1. import the raw phosphorus growth data, and  
2. stardardize the data for each strain by the sum of growth rates.

```{r}


```

### B. Trait Manipulations

In the R code chunk below, do the following:  
1. calculate the maximum growth rate ($\mu_{max}$) of each isolate across all phosphorus types,  
2. create a function that calculates niche breadth (*nb*), and  
3. use this function to calcuate *nb* for each isolate.

```{r}


```  

### C. Visualzing Traits on Trees

In the R code chunk below, do the following:  
1. pick your favorite substitution model and make a Neighbor Joining tree,  
2. define your outgroup and root the tree, and  
3. remove the outgroup branch.

```{r}


```

In the R code chunk below, do the following:  
1. define a color palette (use something other than "YlOrRd"),  
2. map the phosphorus traits onto your phylogeny,  
3. map the *nb* trait on to your phylogeny, and  
4. customize the plots as desired (use `help(table.phylo4d)` to learn about the options).


```{r}


```

***Question 6***:  Based on the distribution of traits on the neighbor joining tree that we created, what might you predict about the degree of specialization of bacteria on diverse phosphorus resources?    
**I don't know that we've provided them with enough information in the handout for them how to interpret this output: what kind of patterns should they look for that would suggest strong or weak phylogenetic signal? how meaningful is variation in std growth ranging from 0.0 - 0.6? likewise, what is the possible range of niche breadth values and how do these two figures relate back to the  generalist--specialist trade-off? Perhaps that would be a good question for them to answer: "formulate a hypothesis that would lend support for the generalist-specialist trade-off, and what kind of patterns would you expect to see from growth rate & niche breadth values that would support this hypothesis?" - NW <-- I totally agree with this --KL**

> ***Answer 6***:   


## 6) HYPOTHESIS TESTING
### A) Phylogenetic Signal: Pagel's Lambda 

In the R code chunk below, do the following:  
1. create two rescaled phylogenetic trees using lambda values of 0.5 and 0,   
2. plot your original tree and the two scaled trees, and  
3. label and customize the trees as desired.

```{r}


```

In the R code chunk below, do the following:  
1. use the `fitContinuous()` function to compare your original tree to the transformed trees.

```{r}


```

***Question 7***:  There are two important outputs from the `fitContinuous()` function that can help you interpret the phylogenetic signal in trait data sets. 
a. First, compare the lambda values of the untransformed tree to the transformed (lambda = 0).
b. Second, compare the Akaike information criterion (AIC) scores of the two models. Differences in AIC scores provide maximum likelihood inference regarding the quality of model fit to a given data set. Models with lower AIC values are better. However, if the difference in AIC values between two models ($\Delta$ AIC) is not greater than at least 2, then then the models are considered equivalent. 
With this information in hand, what does Pagel's lambda say about the phylogenetic signal of niche breadth in our data set? **<-- With the information provided, the students are going to tell you the right answer but have no idea what it actually means, lol**

> ***Answer 7a***:  
> ***Answer 7b***:  


### B) Phylogenetic Signal: Blomberg's K 

In the R code chunk below, do the following:  
1. correct tree branch-lengths to fix any zeros,  
2. calculate Blomberg's K for each phosphorus resource using the `phylosignal()` function,  
3. use the Benjamini-Hochberg method to correct for false discovery rate, and  
4. calculate Blomberg's K for niche breadth using the `phylosignal()` function.

```{r}


```

***Question 8***: Using the K-values and associated p-values (i.e., "PIC.var.P"") from the `phylosignal` output, answer the following questions:

a.  Is there significant phylogenetic signal for niche breadth or standardized growth on any of the phosphorus resources?  
b.  If there is significant phylogenetic signal, are the results suggestive of clustering or overdispersion?  

> ***Answer 8a***:   
> ***Answer 8b***:   


### C.  Calculate Dispersion of a Trait

In the R code chunk below, do the following:  
1. turn the continuous growth data into categorical data,  
2. add a column to the data with the isolate name,  
3. combine the tree and trait data using the `comparative.data()` function in `caper`, and  
4. use `phylo.d()` to calculate *D* on at least three phosphorus traits.

```{r}


```

***Question 9***: Using the estimates for *D* and the probabilities of each phylogenetic model, answer the following questions:

a.  Choose three phosphorus growth traits and test whether they are significantly clustered or overdispersed?  
b.  How do these results compare the results from the Blomberg's K analysis?  
c.  Discuss what factors might give rise to differences between the metrics.  

> ***Answer 9a***:  
> ***Answer 9b***:  
> ***Answer 9c***:  



## 7) PHYLOGENETIC REGRESSION

In the R code chunk below, do the following:  
1. Load and clean the mammal phylogeny and trait dataset,
2. Fit a linear model to the trait dataset, examining the relationship between mass and BMR,
2. Fit a phylogenetic regression to the trait dataset, taking into account the mammal supertree



```{r}


```
  

***Question 10***:


## 7) SYNTHESIS

Below is the output of a multiple regression model depicting the relationship between the maximum growth rate ($\mu_{max}$) of each bacterial isolate and the niche breadth of that isolate on the 18 different sources of phosphorus. 
One feature of the study which we did not reveal earlier in the handout is that the isolates came from two different lakes.  <- this information is in the handout now I think
One of the lakes is an very oligotrophic (i.e., low phosphorus) ecosystem named Little Long (LL) Lake. 
The other lake is an extremely eutrophic (i.e., high phosphorus) ecosystem named Wintergreen (WG) Lake.
We included a "dummy variable" (D) in the multiple regression model (0 = WG, 1 = LL) to account for the environment from which the bacteria were obtained.  

```{r, echo=FALSE, fig.width=6, fig.height=4}
p.growth <- read.table("./data/p.isolates.raw.growth.txt", sep = "\t", header = TRUE, row.names = 1)
umax <- (apply(p.growth, 1, max)) # calculate max growth
lake <- ifelse(grepl("WG",row.names(p.growth)),'WG', 'LL') # make an empty vector for lake id
tradeoff <- data.frame(nb,umax,lake) # make new data frame

D <- (lake == "LL") * 1
fit<-lm(log10(umax) ~ nb + D + nb * D)

LL.nb <- subset(nb,lake == "LL")
WG.nb <- subset(nb,lake == "WG")

LL.umax <- subset(umax,lake == "LL")
WG.umax<-subset(umax,lake == "WG")
par(mar=c(4,4,0,1)+ 0.1)
plot(LL.nb, log10(LL.umax), axes = F, xlab = "Niche Breadth", ylab = "Maximum Growth Rate", 
  pch = 21, cex = 2.0, las = 1, col = "black", bg = "white",
  xlim=c(0,1), ylim = c(-2, 1))
  
  points(WG.nb,log10(WG.umax),pch=21,cex=2,col="black", bg="black")
  box()
  ticks <- c(0.01, 0.1, 1, 10)
  axis(side = 1, labels = T, cex.axis = 1)
  axis(side = 2, las = 1, cex.axis = 1, labels = ticks,
    at = log10(ticks))
  axis(3,labels = F)
  axis(side = 4, at = log10(ticks), labels = F)

curve(fit$coefficients[1] + fit$coefficients[2] * x, from = min(WG.nb - 0.1), 
  to = max(WG.nb +0.1), add = TRUE, lty = 2)
Int <- fit$coefficients[1] + fit$coefficients[3]
Slp <- fit$coefficients[2] + fit$coefficients[4]
curve((Int) + (Slp) * x, from = min(LL.nb - 0.08), to =max(LL.nb + 0.1), add = TRUE, lty = 2)
  
  legend("topleft", legend = c("LL","WG"), pch = c(1,16), cex = 1, col = "black", bty = "n")

summary(fit)

coefficients <- coefficients(fit) # model coefficients
confit <- confint(fit,level=0.95) # 95% CI for parameters
predicted <- fitted(fit) # predicted values
residuals <- residuals(fit) # residuals
anova <- anova(fit) # anova table
vcov <- vcov(fit) # covariance matrix for model parameters
influence <- influence(fit) # regression diagnostics
par(mfrow = c(2, 2), mar = c(5.1 ,4.1 ,4.1 ,2.1))
plot(fit)
```


> we just gave them a bunch of code above, but they can answer the following question without really knowing much about what's going on. Is there a way we can reframe this question (which I think is really important, since it's the culmination of the question laid out at the beginning of the assignment) to be more interactive? Have them code some of it up so they have to think about how niche breadth relates to growth rates and how to use these in a regression? 

***Question 11***: Based on your knowledge of the traits and their phylogenetic distributions, what conclusions would you draw about our data and the evidence for a generalist-specialist tradeoff? 


> ***Answer 11***:


