---
title: 'Handout: Temporal Diversity'
author: 'Z620: Quantitative Biodiversity, Indiana University'
date: "February 10, 2017"
header-includes:
   - \usepackage{array}
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

In this exercise, we will explore aspects of temporal diversity. We will first introduce you to the time-by-site-species matrix and how to manage this type of ecological data structure. We will then...

We will review ecological concepts such as community turnover/ temporal beta diversity and stability as well as statistical concepts such as autocorrelation. In addition, we will review statistical tools such as repeated measures ANOVA, autocorrelation-corrected regressionas well as the visualization tool of the Muller plot. 

## 1) SETUP
### Retrieve and Set Your Working Directory

```{r, results = 'hide'}
rm(list=ls()) 
getwd() 
#setwd("~/GitHub/QuantitativeBiodiversity/QB-2017/Week5-Temporal/") 
```

### Install Packages

```{r, results = 'hide', message = FALSE, warning = FALSE} 
package.list <- c('vegan', 'tidyr', 'dplyr', 'codyn', 'ggplot', 'cowplot')
for (package in package.list) {
  if (!require(package, character.only=T, quietly=T)) {
    install.packages(package)
    library(package, character.only=T)
  }
}
```

## 2) LOADING DATA

We will start by using the long-term monitoring dataset of rodents from the Chihuahuan Desert ecosystem near Portal, Arizona.

```{r}
portal <- read.table("data/combined.csv", sep = ",", header = TRUE)
```

## 3) MANAGING AND MANIPULATING DATA
```{r}
# combine dates into one column
portal <- unite(portal, col = date, c(year, month, day), sep = "-", remove = FALSE)

# combine species into one column
portal <- unite(portal, col = taxon, c(genus, species), sep = "_", remove = FALSE)

# Create the time-by-species matrix
time.by.species <- group_by(portal, date) %>% count(taxon) %>% spread(key = taxon, value = n, fill = 0)
time.by.species <- as.data.frame(time.by.species)
head(time.by.species)

# make column 1 the row names
rownames(time.by.species) <- time.by.species[,1]
time.by.species <- time.by.species[,-c(1,2)]
head(time.by.species)

dim(time.by.species)

```

## 4) BASIC TIME SERIES ANALYSIS
-- single population
-- identifying and correcting autocorrelation
-- perhaps something from the following:
http://www.statmethods.net/advstats/timeseries.html
https://a-little-book-of-r-for-time-series.readthedocs.io/en/latest/
https://cran.r-project.org/web/views/TimeSeries.html
https://www.analyticsvidhya.com/blog/2015/12/complete-tutorial-time-series-modeling/
http://www.stat.pitt.edu/stoffer/tsa4/R_toot.htm
http://www.statoek.wiso.uni-goettingen.de/veranstaltungen/zeitreihen/sommer03/ts_r_intro.pdf

## 5) REPEATED MEASURES ANOVA
-- code from Megan
-- for univariate repsonse variable
-- perhaps revisit a diversity metric from week 2

## 6) TEMPORAL BETA DIVERSITY

```{r}
portal.richness.year <- group_by(portal, year) %>% count(taxon) %>% summarise(n())
rich.plot <- ggplot(portal.richness.year, aes(x = year, y = `n()`, facet(`n()`))) +
  geom_line(size = 1) + 
  xlim(1977, 2002) + 
  xlab("Year") + 
  ylab("Richness")

plot(rich.plot)

portal.species.abunds <- group_by(portal, year) %>% count(taxon)
portal.turnover <- turnover(df = portal.species.abunds, 
                            time.var = "year", 
                            species.var = "taxon", 
                            abundance.var = "n",
                            metric = "total")

portal.appearance <- turnover(df = portal.species.abunds, 
                            time.var = "year", 
                            species.var = "taxon", 
                            abundance.var = "n",
                            metric = "appearance")

portal.disappearance <- turnover(df = portal.species.abunds, 
                            time.var = "year", 
                            species.var = "taxon", 
                            abundance.var = "n",
                            metric = "disappearance")
portal.turnover$metric<-"total"
names(portal.turnover)[1]="turnover"
  
portal.appearance$metric<-"appearance"
names(portal.appearance)[1]="turnover"
  
portal.disappearance$metric<-"disappearance"
names(portal.disappearance)[1]="turnover"
  
portal.allturnover<-rbind(portal.turnover, portal.appearance, portal.disappearance)
names(portal.allturnover)[2] <- "year"

turn.plot <- ggplot(portal.allturnover, aes(x = year, y = turnover, color = metric)) +
  geom_line(size = 1, show.legend = F) + 
  xlim(1977, 2002) + 
  xlab("Year") + 
  ylab("Turnover")
plot(turn.plot)



portal.rankshift <- rank_shift(df = portal.species.abunds, 
                               time.var = "year",
                               species.var = "taxon",
                               abundance.var = "n")
portal.rankshift$year <- as.numeric(substr(portal.rankshift$year_pair, 6, 9))

rankshift.plot <- ggplot(portal.rankshift, aes(x = year, y = MRS)) + 
  geom_line(size = 1) + 
  xlim(1977, 2002) + 
  xlab("Year") + 
  ylab("Mean Rank Shift")



# Check out help(rate_change_interval) and note that this function calculates the 
# Euclidean distance between each site. Remember, this is not an ideal distance.
# Let's use an appropriate transformation on the abundances first.

# let's cound the total abundances in each year (i.e., rowsum)
total.abunds <- portal.species.abunds %>% group_by(year) %>% count(wt = n)

# Now, let's join these total counts with the counts for each species
portal.hellinger.transf <- inner_join(portal.species.abunds, total.abunds, by = "year") %>% 
  mutate(hellinger.transf = sqrt(n / nn))

# Now, we have a third column, the square root of the relative abundances, the hellinger transformation
# calculating the euclidean distance on Hellinger-transformed data is appropriate.
portal.change.int <- rate_change_interval(portal.hellinger.transf,
                     time.var = "year",
                     species.var = "taxon",
                     abundance.var = "hellinger.transf")

rate.plot <- ggplot(portal.change.int, aes(interval, distance)) + 
  geom_point() + 
  stat_smooth(method = "loess", se = F, size = 1) + 
  ylab("Hellinger Distance") + 
  xlab("Time Interval (Years)")
rate.plot

turnover.grid.plot <- plot_grid(rich.plot, 
          turn.plot, 
          rankshift.plot,
          rate.plot,
          labels = c("A", "B", "C", "D"), 
          nrow = 4, align = "v")

ggsave("fig-turnover-grid.png", turnover.grid.plot,
       width = 6, height = 10, units = "in")
graphics.off()
grid::grid.raster(png::readPNG("fig-turnover-grid.png"))

```


## 7) SYNCHRONY, or covarying species; perhaps stability, compensatory dynamics, or variance ratio?
http://onlinelibrary.wiley.com/doi/10.1111/2041-210X.12188/epdf
https://cran.r-project.org/web/packages/vrtest/vrtest.pdf

## 8) Using environmental drivers to explain temporal biodiversity data (univariate, multivariate, both?)