---
title: 'Handout: Temporal Diversity'
author: 'Z620: Quantitative Biodiversity, Indiana University'
date: "February 10, 2017"
header-includes:
   - \usepackage{array}
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

In this exercise, we will explore aspects of temporal diversity. We will first introduce you to the time-by-site-species matrix and how to manage this type of ecological data structure. We will then...

We will review ecological concepts such as community turnover/ temporal beta diversity and stability as well as statistical concepts such as autocorrelation. In addition, we will review statistical tools such as repeated measures ANOVA, autocorrelation-corrected regressionas well as the visualization tool of the Muller plot. 

## 1) SETUP
### Retrieve and Set Your Working Directory

```{r, results = 'hide'}
rm(list=ls()) 
getwd() 
setwd("~/GitHub/QB-2017/Week5-Temporal/") 
```

### Install Packages

```{r, results = 'hide', message = FALSE, warning = FALSE} 
package.list <- c('vegan', 'tidyr', 'dplyr', 'codyn', 'ggplot2', 'cowplot', 'MullerPlot', 'RColorBrewer', 'reshape2', 'lubridate')
for (package in package.list) {
  if (!require(package, character.only=T, quietly=T)) {
    install.packages(package)
    library(package, character.only=T)
  }
}
```

## 2) LOADING DATA

We will start by using the long-term monitoring dataset of rodents from the Chihuahuan Desert ecosystem near Portal, Arizona.
The so-called "Portal Project" <http://portal.weecology.org/> was initiatied in the late 1970s by Jim Brown and colleagues to look at species interactions, specifically competition between rodents and ants for plant seeds.
Initial experiments focused on removing rodents with fencing and trapping, which led to an initial increase in the ant population size as well as changes in plant species composition.
These promising findings inspired Brown and colleagues to construct an improved experiment design that continues today. 
The Portal Project monitors 24 replicated experimental plots, each 0.25 ha in area (50 X 50 m).
In total, 4.8 kilometers of fencing is used in the Portal experiments!

Let's take a look at the data:

```{r}
portal <- read.table("~/GitHub/QB-2017/Week5-Temporal/data/combined.csv", sep = ",", header = TRUE)
```

There are a few things to point out about the data. 
First, the version of the data that we are working with spans from 1977 - 2002. 
During this period, individual rodents were captured from plots ("plot_id"), identified to species ("species_id"), which led to the creation of a "record_id" with an associated date (day, month, and year).
In addition, the sex, size ("hindfoot_length" and "weight") and taxonomic identity ("genus" and "species") of each animal was recorded. 
All of this was done in five experimentally replicated treatments. 

1) Controls - fencing around plots does not exclude rodents 
2) Long-term Krat - long-term exclusion of Kangaroo rats (*Dipodomys merriami*)
3) Short-term Krat - short-term exclusion of Kangaroo rats (*Dipodomys merriami*)
4) Rodent Exclosure - exclusions of all rodents
5) Spectab exclosure - exclusion of Banner-tailed kangaroo rat (*Dipodomys spectabilis*) 

## 3) MANAGING AND MANIPULATING DATA

The Portal Project is a great example of a long-term biodiversity project. 
In order for such a project to be successful, the scientiss need to not only generate a lot of data, but make sure it is well organized. 
One thing you may notice when looking at `portal` is that it has a lot of observations, almost 35,000.
These data are entered in the *long format*, which means that, as described above, each row represents a unique observation. 
The long format is a convenient way to enter data that reduces entry errors.
However, you might need to manipulate the data for certain analyses. 
For example, many R packages and functions will want you to organize your data in *wide format* where different sampling time points would be represented in columns. 
In the following section, we will briefly demonstate how to manipulate the `portal` data set to carry out different types of analyses using function from the `dplyr` package, which was designed for transforming and summarizing tabulard data. 

```{r}
# Here, we're combining dates into one column
portal <- unite(portal, col = date, c(month, day, year), sep = "/", remove = FALSE)
portal$date <- as.Date(portal$date, "%m/%d/%Y")

# Now let's combine taxonomic data into one column
portal <- unite(portal, col = taxon, c(genus, species), sep = "_", remove = FALSE)

# Create the time-by-species matrix
time.by.species <- group_by(portal, date) %>% count(taxon) %>% spread(key = taxon, value = n, fill = 0)
time.by.species <- as.data.frame(time.by.species)
head(time.by.species)

# make column 1 the row names
rownames(time.by.species) <- time.by.species[,1]
time.by.species <- time.by.species[,-c(1,2)]
head(time.by.species)

dim(time.by.species)

```


## 4) VISUALIZING TIME SERIES DATA 

```{r}
remove <- c("2000-4-31", "2000-9-31")
time.by.species <- time.by.species[!rownames(time.by.species) %in% remove, ]

time.by.species.T <- t(time.by.species)
species <- rownames(time.by.species.T)
parents <- rep(NA, each=length(species))

n <- length(species)
qual_col_pals <- brewer.pal.info[brewer.pal.info$category == 'qual',]
col_vector <- unlist(mapply(brewer.pal, qual_col_pals$maxcolors, rownames(qual_col_pals)))
col <- sample(col_vector, n)

attributes.portal <- cbind(species, parents, col) 
class(time.by.species.T) <- "numeric" 
long.form <- melt(time.by.species.T)
dates <- as.vector(long.form$Var2)
oldest <- as.Date("1977-7-16")

long.form$Var4 <- as.Date(as.character(long.form$Var2)) - oldest

keeps <- c("Var1", "Var4", 'value')
long.form <- long.form[keeps]

m.plot <- Muller.plot(attributes = attributes.portal, population.data = long.form,
                      data.method = "list", time.interval.method = "linear")

plot(0,0,axes = F)
par(mar=c(0, 0, 0, 0))
legend("right", legend = m.plot$name,col = as.character(m.plot$color),pch = 19)


```


## 4) BASIC TIME SERIES ANALYSIS
```{r}
# Let's choose a control plot
port.cont <- filter(portal, plot_id == 2)

# Now let's sum the mass of all rodents collected on a given date
summary <- port.cont %>%
  group_by(date) %>%
  summarize(rod.mass = sum(weight), na.rm = TRUE)

# Let's take a look at the plot of rodent mass over time in the plot
plot(rod.mass ~ date, summary, xaxt = "n", type = "l")
```

-- single population
-- identifying and correcting autocorrelation
-- perhaps something from the following:
http://www.statmethods.net/advstats/timeseries.html
https://a-little-book-of-r-for-time-series.readthedocs.io/en/latest/
https://cran.r-project.org/web/views/TimeSeries.html
https://www.analyticsvidhya.com/blog/2015/12/complete-tutorial-time-series-modeling/
http://www.stat.pitt.edu/stoffer/tsa4/R_toot.htm
http://www.statoek.wiso.uni-goettingen.de/veranstaltungen/zeitreihen/sommer03/ts_r_intro.pdf

## 5) REPEATED MEASURES ANOVA
-- code from Megan
-- for univariate repsonse variable
-- perhaps revisit a diversity metric from week 2

## 6) TEMPORAL BETA DIVERSITY

```{r}
portal.richness.year <- group_by(portal, year) %>% count(taxon) %>% summarise(n())
rich.plot <- ggplot(portal.richness.year, aes(x = year, y = `n()`, facet(`n()`))) +
  geom_line(size = 1) + 
  xlim(1977, 2002) + 
  xlab("Year") + 
  ylab("Richness")

plot(rich.plot)

portal.species.abunds <- group_by(portal, year) %>% count(taxon)
portal.turnover <- turnover(df = portal.species.abunds, 
                            time.var = "year", 
                            species.var = "taxon", 
                            abundance.var = "n",
                            metric = "total")

portal.appearance <- turnover(df = portal.species.abunds, 
                            time.var = "year", 
                            species.var = "taxon", 
                            abundance.var = "n",
                            metric = "appearance")

portal.disappearance <- turnover(df = portal.species.abunds, 
                            time.var = "year", 
                            species.var = "taxon", 
                            abundance.var = "n",
                            metric = "disappearance")
portal.turnover$metric<-"total"
names(portal.turnover)[1]="turnover"
  
portal.appearance$metric<-"appearance"
names(portal.appearance)[1]="turnover"
  
portal.disappearance$metric<-"disappearance"
names(portal.disappearance)[1]="turnover"
  
portal.allturnover<-rbind(portal.turnover, portal.appearance, portal.disappearance)
names(portal.allturnover)[2] <- "year"

turn.plot <- ggplot(portal.allturnover, aes(x = year, y = turnover, color = metric)) +
  geom_line(size = 1, show.legend = F) + 
  xlim(1977, 2002) + 
  xlab("Year") + 
  ylab("Turnover")
plot(turn.plot)



portal.rankshift <- rank_shift(df = portal.species.abunds, 
                               time.var = "year",
                               species.var = "taxon",
                               abundance.var = "n")
portal.rankshift$year <- as.numeric(substr(portal.rankshift$year_pair, 6, 9))

rankshift.plot <- ggplot(portal.rankshift, aes(x = year, y = MRS)) + 
  geom_line(size = 1) + 
  xlim(1977, 2002) + 
  xlab("Year") + 
  ylab("Mean Rank Shift")



# Check out help(rate_change_interval) and note that this function calculates the 
# Euclidean distance between each site. Remember, this is not an ideal distance.
# Let's use an appropriate transformation on the abundances first.

# let's cound the total abundances in each year (i.e., rowsum)
total.abunds <- portal.species.abunds %>% group_by(year) %>% count(wt = n)

# Now, let's join these total counts with the counts for each species
portal.hellinger.transf <- inner_join(portal.species.abunds, total.abunds, by = "year") %>% 
  mutate(hellinger.transf = sqrt(n / nn))

# Now, we have a third column, the square root of the relative abundances, the hellinger transformation
# calculating the euclidean distance on Hellinger-transformed data is appropriate.
portal.change.int <- rate_change_interval(portal.hellinger.transf,
                     time.var = "year",
                     species.var = "taxon",
                     abundance.var = "hellinger.transf")

rate.plot <- ggplot(portal.change.int, aes(interval, distance)) + 
  geom_point() + 
  stat_smooth(method = "loess", se = F, size = 1) + 
  ylab("Hellinger Distance") + 
  xlab("Time Interval (Years)")
rate.plot

turnover.grid.plot <- plot_grid(rich.plot, 
          turn.plot, 
          rankshift.plot,
          rate.plot,
          labels = c("A", "B", "C", "D"), 
          nrow = 4, align = "v")

ggsave("fig-turnover-grid.png", turnover.grid.plot,
       width = 6, height = 10, units = "in")
graphics.off()
grid::grid.raster(png::readPNG("fig-turnover-grid.png"))

```


## 7) SYNCHRONY, or covarying species; perhaps stability, compensatory dynamics, or variance ratio?
http://onlinelibrary.wiley.com/doi/10.1111/2041-210X.12188/epdf
https://cran.r-project.org/web/packages/vrtest/vrtest.pdf

## 8) Using environmental drivers to explain temporal biodiversity data (univariate, multivariate, both?)