---
title: 'Handout: Temporal Diversity'
author: 'Z620: Quantitative Biodiversity, Indiana University'
date: "February 10, 2017"
header-includes:
   - \usepackage{array}
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

In this exercise, we will explore aspects of temporal diversity. We will review ecological concepts such as community turnover/ temporal beta diversity and stability as well as statistical concepts such as autocorrelation. In addition, we will review statistical tools such as repeated measures ANOVA, autocorrelation-corrected regressionas well as the visualization tool of the Muller plot. 

## 1) SETUP
### Retrieve and Set Your Working Directory

```{r, results = 'hide'}
rm(list=ls()) 
getwd() 
setwd("~/GitHub/QB-2017/Week5-Temporal") 
```

### Install Packages

```{r, results = 'hide', warning=FALSE, message=FALSE}

```

## 2) LOADING DATA

We will start by using the long-term monitoring dataset of rodents from the Chihuahuan Desert ecosystem near Portal, Arizona.

```{r}
portal <- read.table("data/Portal_rodents_19772002.csv", sep = ",", header = TRUE)
```

## 3) VISUALIZING TURNOVER

First, we will write a function to clean the data and generate a time-by-species matrix. 

```{r}
portal.dates <- portal[,2:4]
portal.dates.unique <- unique(portal.dates)

# To-do. Initialize an empty dataframe with columns for day, month, year, and each species

f <- function(x) {
 new.data <- portal[ which( portal$mo == x[1] & portal$dy == x[2] & portal$yr == x[3]) , ]
 species <- array(new.data$species)
 remove <- c("<NA>")
 species <- species[! species %in% remove]
 species.df <- as.data.frame(table(species))
 # append species abundances to empty dataframe, making sure to match species identity to column
 #print(as.data.frame(table(species)))
}

test <- apply(portal.dates.unique, 1, f)
```

