---
title: "Assignment: Temporal Diversity"
author: "Student Name; Z620: Quantitative Biodiversity, Indiana University"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

In this Assignment, we extend our understanding of diversity from the spatial dimension to the temporal dimension. 

After completing this exercise you will know how to:

1. wrangle a large dataset to visualize and analyze time series data
2. test hypotheses from experiments with temporal data
3. quantify temporal $\beta$-diversity and stability

## Directions:
1. Change "Student Name" on line 3 (above) with your name.
2. Complete as much of the exercise as possible during class; what you do not complete in class will need to be done on your own outside of class.
3. Use the Handout as a guide; it contains a more complete description of data sets along with the proper scripting needed to carry out the exercise.
4. Be sure to **answer the questions** in this exercise document; they also correspond to the Handout.
Space for your answer is provided in this document and indicated by the ">" character.
If you need a second paragraph be sure to start the first line with ">".
5. Before you leave the classroom, **push** this file to your GitHub repo.
6. When you are done wit the Assignment, **Knit** the text and code into a html file.
7. After Knitting, please submit the completed Assignment by creating a **pull request** via GitHub.
Your pull request should include this file *temporal_assignment.Rmd* and the html output of `Knitr` (*temporal_assignment.html*).

## 1) R SETUP

Typically, the first thing you will do in either an R script or an RMarkdown file is setup your environment. 
This includes things such as setting the working directory and loading any packages that you will need.

In the R code chunk below, provide the code to:  

1. clear your R environment,
2. print your current working directory,
3. set your working directory to your "*/Week5-Temporal*" folder, and
4. load the `vegan` R package (be sure to install if needed).  

```{r}




```

## 2) LOADING DATA
### Load dataset
In the R code chunk below, do the following:  

1. load the `portal` dataset from in the "*/Week5/data*" folder, and
2. explore the structure of the dataset.

```{r}




```

***Question 1***:  Describe some of the attributes of the `portal` dataset.  

a.  How many plots are in `portal`?
b.  How many rodent species are there in the `portal` dataset?

> ***Answer 1a***:  
> ***Answer 1b***:  


## 3) WRANGLING THE PORTAL DATASET

In the R code chunk below, do the following:  

1. Create a site-by-species matrix for any year of your choosing.
2. Create a vector of plot_type for sites in the site-by-species matrix.
3. Calculate alpha diversity across the sites for that year.
4. Using the hypothesis testing tools you learned in the beta-diversity module, test the hypothesis that species abundances across sites vary as a factor of treatment type (i.e., plot_type). 

```{r}

```

***Question ***: Does richness differ among 

> ***Answer ***:




## 4) TIME SERIES ANALYSIS
In the R code chunk below, do the following:  

1. 
2. 

```{r}


```

***Question ***:

> ***Answer ***:



## 5) TURNOVER
In the R code chunk below, do the following:  

1. 
2. 

```{r}


```

***Question ***:

> ***Answer ***:


## 6) STABILITY
In the R code chunk below, do the following:  

1. 
2. 

```{r}


```

***Question ***:

> ***Answer ***:



## SYNTHESIS
In the R code chunk below, do the following:  

1. 
2. 

```{r}


```

***Question ***:

> ***Answer ***:

