---
title: "Among Site (Beta) Diversity"
author: "Z620: Quantitative Biodiversity, Indiana University"
date: "February 6, 2015"
header-includes:
   - \usepackage{array}
output: pdf_document
geometry: margin=2.54cm
---

##Overview##
Following R chunk take data file with lots of taxonomy information (characters) and OTU data (site-by-species matrix) and turns it into a list. 
This file creates the RData file "soilbac.RData" used in the Beta exercise. 
This script was run by the instructors prior to class and the output was saved in the data folder.

## 1) SETUP
### A. Retrieve and Set Your Working Directory
```{r, results = 'hide'}
rm(list = ls())
getwd()
setwd("~/GitHub/QuantitativeBiodiversity/Assignments/Beta")
```

## 2) LOAD DATA
```{r}
soilbac.full <- read.table("./data/soilbacfull.txt", sep = "\t", header = TRUE, row.names = 1)
```

##3) CREATE LIST
```{r}
tax <- data.frame(soilbac.full[,1:5]) #assign taxonomy
otu <- t(data.frame(soilbac.full[,6:28])) #assign site-by-species matrix
soilbac <- list(otu, tax) # create list
names(soilbac) <- c("otu", "tax") # name objects in list
str(soilbac, max.level = 1) # look at structure of list
```

##4) SAVE LIST 
```{r}
save(soilbac, file = "./data/soilbac.RData")
```

##5) LOAD LIST
Somethig is not quite right here, probably because I'm not writing the table correctly.
```{r}
rm(list = ls())
load("./data/soilbac.RData")
```
