---
title: "Add title of your project"
author: "Add names of team members, Quantitative Biodiversity, Indiana University"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
geometry: margin=2.54cm
---

## Directions:
1. Change "Add title of your project" on line 2 (above) to the name of your project.
2. Change "Add names of team members" on line 3 (above) to the names of team members in your project.
3. Write a <250 word Abstract describing your team's research.
4. Write a 1-2 paragraph Introduction providing some background and context for your project.
5. Set up your working enviroment (set directory, load packages) using the `/Week8-Projects` folder. 
6. Provide a brief (~1 paragraph) description of the dataset.
7. Load the dataset and process in ways necessary for any initial analyses.
8. Use well annotated R chunks to analyze data, make figures, and peform statistics needed to address your quesstions and hypotheses. 
9. Write a Discussion and Conclusion section to address the questions and hypotheses that you outlined in you outlined in your Abstract and Introuction.
10. Include citations in the Reference section.
11. Delete the text in this Directions section.
12. By Thursday March 9, 23:59, please submit the completed assignment by creating a **pull request** via GitHub. Your pull request should include this file *Project_assignment.Rmd* and the PDF output of `Knitr` (*Project_assignment.pdf*). 

## ABSTRACT

## 1) INTRODUCTION

## 2) SETUP

```{r}

```

## 3) DESCRIPTION OF DATA

## 4) LOAD THE DATA

```{r}


```

## 5) ANALYSIS: FIGURES AND STATISTICS

```{r}


```

## 6) DISCUSSION AND CONCLUSION

## 7) REFERENCES
